---
title: "Assignment 3 Spatial Epidemiology"
author: "Ferrara Lorenzo, Lucchini Marco"
# date: "25-10-2022"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r include=FALSE}
# knitr::opts_chunk$set( echo=F )
knitr::opts_chunk$set( cache = T )
knitr::opts_chunk$set( warning=FALSE )
knitr::opts_chunk$set( out.width = '60%' )
knitr::opts_chunk$set( fig.align = "center" )
knitr::opts_chunk$set( tidy =TRUE )
knitr::opts_chunk$set( message = FALSE )
```

# Task 1: Introduction to Spatial Point Process

### The nest data from islet “nucli 23” is stored in nucli23.txt. Additionally, the coordinates of the islet are in poly23.txt.

### 1) Build a ppp object using the “poly23” data as a window

```{r include=FALSE}
rm(list=ls())
library(spatstat)
```

```{r warning=FALSE}
#the data
nucli.23=read.delim("data/T1/nucli23.txt")
min.X = min(nucli.23$X)
min.Y = min(nucli.23$Y)

nucli.23$X=nucli.23$X-min.X
nucli.23$Y=nucli.23$Y-min.Y

max.X = max(nucli.23$X)
max.Y = max(nucli.23$Y)
```

```{R}
#the polygon object for the border
poligon=poly23 <- read.delim("data/T1/poly23.txt")
poligon$X=poligon$X-min.X
poligon$Y=poligon$Y-min.Y
pol.illa<-list(x=poligon$X,y=poligon$Y)

min.pX = min(poligon$X)
min.pY = min(poligon$Y)
max.pX = max(poligon$X)
max.pY = max(poligon$Y)
```

```{R}
#the final object
n23p=ppp(nucli.23$X,nucli.23$Y,poly=pol.illa)
par(mfrow=c(1,1), font = 2, font.axis = 2, font.lab = 4, las = 1, mar = c(1, 0, 1, 0))
plot(n23p,main="Nests")
axis(1, at = c(floor(seq(min.pX, max.pX, by=25))), pos=c(min.pX,min.pY-15))
axis(2, at = c(floor(seq(min.pY, max.pY, by=25))), pos=c(-10,0))
```

### 2) Describe the point pattern process and its intensity.

```{r}
summary(n23p)
```
```{r}
par(mfrow=c(1,1), font = 2, font.axis = 2, font.lab = 4, las = 1, mar = c(0, 0, 2, 0))
plot(density(n23p, dimax.Yx=c(256,256), sigma=5.5), main="Intensity of the point process")
```

The density plot reveals that the majority of the nests are located in the center and south, along the north coast. There is a notable concentration, indicated by the yellow peak, in which a large number of nests were found. 

This distribution does not appear to be completely random and appears to have a pattern.

### 3) Create a multi-type mark indicating the order of the nesting according to the nesting time, Using the time intervals: [10,16], [17,19], [20,22].

```{r}
n23T=ppp(nucli.23$X,nucli.23$Y,poly=pol.illa,marks=nucli.23$data_pos)
```


```{r}
par(mfrow=c(1,1), font = 2, font.axis = 2, font.lab = 4, las = 1, mar = c(2, 0, 2, 0))
plot(n23T,main="Nests",markscale=0.3,leg.side="right")
axis(1,at=c(seq(0,max.pX, by=20)),pos=c(-10,0))
axis(2,at=c(seq(-10,max.pY,by=25)),pos=c(-10,0))
```

```{r}
DPOScat=cut(nucli.23$data_pos, breaks=c(9, 16, 19, 22), labels=c("10-16","17-19","20-22"))
table(DPOScat)
```


The suddivision of the nests in the 3 temporal groups seems homogeneous

```{r}
n23Tcat=ppp(nucli.23$X,nucli.23$Y,poly=pol.illa,marks=DPOScat)
summary(n23Tcat)
```

### 4) Describe the marked point process and its intensity

```{r}
par(mfrow=c(1,1), font = 2, font.axis = 2, font.lab = 4, las = 1, mar = c(0, 0, 1, 0))
plot(split(n23Tcat), main="Nests grouped by nesting time")
```

```{r}
par(mfrow=c(1,1), font = 2, font.axis = 2, font.lab = 4, las = 1, mar = c(0, 0, 1, 0))
plot(density(split(n23Tcat),sigma=10), ribbon = FALSE, main="Intensity plots")
```

Different patterns can be observed in the three time intervals: there is an initial distribution in the north and south of the islet, which gradually fades towards the center.

### 5) Add to the analysis the height of the islet.

```{r}
grid <- read.csv("data/T1/grid_height_23.txt", sep="")
mat<-as.matrix(read.table("data/T1/height_23.txt"))
mat[mat==0]=NA
height<-im(mat,grid$x,grid$y)
par(mfrow=c(1,1), font = 2, font.axis = 2, font.lab = 4, las = 1, mar = c(0, 0, 2, 0))
plot(height,axis=T, main="Comparison between height and nests' position")
plot(n23p, add=T, cex=0.7, pch=16)
```

This plot allows us to observe a correlation between the positions of the nest and the elevation of the islet. It appears that the nests are concentrated in areas with higher elevation. Therefore, we can conclude that the point process is not completely random and the assumption of a Homogeneous Poisson Process can be rejected.


\newpage

# Task 2: Intensity and Randomness

### The nest data from islet “nucli 84” is stored in nucli84.txt. Additionally, the coordinates of the islet are in poly84.txt.

### 1) Build a ppp object using the “nucli 84” data.

```{r include=FALSE}
library(spatstat)
```

```{r}
rm(list=ls())
nucli84 <- read.delim("data/T2/nucli84.txt")

min.X = min(nucli84$X)
min.Y = min(nucli84$Y)
max.X = max(nucli84$X)
max.Y = max(nucli84$Y)

nucli84$X = nucli84$X - min.X
nucli84$Y = nucli84$Y - min.Y

n84=ppp(x = nucli84$X, y = nucli84$Y, range(nucli84$X), range(nucli84$Y))

poligon=read.delim("data/T2/poly84.txt")

poligon$X = poligon$X - min.X
poligon$Y = poligon$Y - min.Y
pol.illa<-list(x=poligon$X,y=poligon$Y)

min.pX = min(poligon$X)
min.pY = min(poligon$Y)
max.pX = max(poligon$X)
max.pY = max(poligon$Y)

n84p=ppp(nucli84$X,nucli84$Y,poly=pol.illa, range(poligon$X), range(poligon$Y))

islet_window = owin(poly = pol.illa)
```

```{r}
par(mfrow=c(1,1), font = 2, font.axis = 2, font.lab = 4, las = 1, mar = c(2, 0, 1, 0))
plot(n84p,main="Nests")
axis(1,at=c(round(seq(min.pX, max.pX,length=10),digits=0)))#, pos=c(0,0))
axis(2,at=c(round(seq(min.pY, max.pY,length=10),digits=0)), pos=c(min.pX-4,min.pY-50))
```

### 2) Draw a plot with the intensity of the point process computed by the non-parametric approach. Briefly comment the results.

```{R}
par(mfrow=c(1,1), font = 2, font.axis = 2, font.lab = 4, las = 1, mar = c(0, 0, 1, 0))
plot(density(n84p, dimax.Yx=c(256,256), sigma=15), main="Intensity of the point process")
```

In the density plot, we can see that the nests are concentrated in two main areas: one in the northwest corner and one in the eastern part, both located on the seaside. No nests were observed in the central or southern areas of the islet.

This may suggest that a geographical feature is influencing the location of the nests.

We also try to fit  a homogeneous Poisson process:

```{r warning=FALSE}
modelPois = ppm(n84, ~ 1)
modelPois
```

So the intensity has point estimate `r exp(modelPois$coef)` and 95% confidence interval 

```{r echo=T, include=F}
exp(confint(modelPois))
```

```{r echo=F}
temp = exp(confint(modelPois))
rownames(temp)=""
temp
```


### 3) Assess the Completely Spatial Randomness hypothesis

- via Chi-square test:

We divide the region in 8 subareas with equal areas and under CSR we would expect more or less same number of nests in each subregion.

```{r}
M <- quadrat.test(n84, nx = 4, ny = 2)
par(mfrow=c(1,1), font = 2, font.axis = 2, font.lab = 4, las = 1, mar = c(0, 0, 0, 0))
plot(n84,main="" )
plot(M, add = TRUE)
```

Nonetheless we notice that some subregions have more nests than other. Indeed after performing a Chi-square test we get:

```{r}
M
```

The p-value is extremely low ($p < 2.2e-16$), which indicates that we can reject the null hypothesis of complete spatial randomness (CRS). 

- via Kolmogorov-Smirnov test:
```{r}
KS=cdf.test(n84,covariate="x")
KS
plot(KS)
```

```{r}
KS=cdf.test(n84,covariate="y")
KS
plot(KS)
```

The observed distribution of Z at data points differs significantly from the expected distribution, allowing us to reject the null hypothesis of CSR and conclude that there is a dependence between the intensity of the points and both the Cartesian coordinates.

### 4) Assess the relation between the intensity of the point process and the covariates height and vegetation.

```{r}
grid <- read.delim("data/T2/grid.txt")
grid.veg = read.delim("data/T2/grid_veg.txt")
veg = as.matrix(read.delim("data/T2/veg.txt",header=FALSE))
height = as.matrix(read.delim("data/T2/height.txt", header=FALSE))
```

First we visualize the values of Height and Vegetation in the islet:

```{R}
par(mfrow=c(1,2), font = 2, font.axis = 2, font.lab = 4, las = 1, mar = c(0, 0, 1, 2))

Height = im(mat = height, xcol = grid$x, yrow = grid$y)
plot(Height, main="Height", clipwin = islet_window)
plot(n84p, add=T, cex=0.5, pch=16)

Vegetation = im(mat = veg, xcol = grid.veg$x, yrow = grid.veg$y)
plot( Vegetation, main="Vegetation", clipwin = islet_window)
plot(n84p, add=T, cex=0.5, pch=16)
```

The nests seem to be concentrated in areas with higher vegetation and elevation.

Now we categorize the covariates using the suggested intervals:

- Height. [0,10], (10,20], (20,40]
- Vegetation. [0,20], (20,50], (50,100]

```{r}
#height
brks.h <- c(0,10,20,40)
Hcut <- cut(Height, breaks = brks.h, labels = c("0-10", "10-20", "20-40"))
H <- tess(image = Hcut)

#vegetation
brks.v <- c(0, 20, 50, 100)
Vcut <- cut(Vegetation, breaks = brks.v, labels = c("0-20", "20-50", "50-100"))
V <- tess(image = Vcut)
```

```{R}
par(mfrow=c(1,2), font = 2, font.axis = 2, font.lab = 4, las = 1, mar = c(0, 0, 1, 3.5))
plot(H, main="Height")
plot(V, main="Vegetation")
```

And to assess quantitatively this relation we use:

- a Chi-squared test:

```{r}
M.h=quadrat.test(n84, tess = H)
M.h

M.v=quadrat.test(n84p, tess = V)
M.v

par(mfrow=c(1,2), font = 2, font.axis = 2, font.lab = 4, las = 1, mar = c(0, 0, 1, 3.5))
plot(M.h, valuesAreColours=FALSE, main="Height", cex=0.6)
plot(M.v, valuesAreColours=FALSE, main="Vegetation", cex=0.6)

```

The Chi-squared tests on the two covariates Height and Vegetation have p-values: $p_1<1 \cdot 10^{-16}$ and $p_2=1.195\cdot 10^{-7}$.

- a Kolmogorov-Smirnov test: 
```{r warning=FALSE}
KS <- cdf.test(n84, Height)
KS

plot(KS,style="QQ")
```

```{r}
KS <- cdf.test(n84, Vegetation)
KS

plot(KS,style="QQ")
```

Also the Kolmogorov test has very little p-values and the plots suggest a general difference of the observed quantiles from the theoretical ones.

So, in the light of the above considerations, we can state there is in fact a relation between the covariates and the point process.

### 5) Fit an inhomogeneous Poisson model to data

We'll fit the model:
$$\lambda \sim x + y + \text{Height} + \text{Vegetation}$$

```{r warning=FALSE}
model = ppm(n84, ~ x + y + Height + Vegetation)
model
```

<!-- We'll do a stepwise feature selection, removing variables that are not significant one at a time. -->

We notice that we can remove the $x$ variable as it's not significant, since it has the Z statistic very close to 0.

<!-- First we remove the $x$ variable as it's not significant and it has the closest Z statistic to 0. -->

```{r warning=FALSE}
model1 = ppm(n84, ~ y + Height + Vegetation)
model1
```

Now all the covariates are significant (they have the Z statistic far enough from zero and their confidence intervals don't contain 0), so we can keep the model:
$$\lambda \sim y + \text{Height} + Vegetation$$

whose fitted parameters have point estimates:
```{r}
chosen.model = model1

coef(chosen.model)
```

and 95% confidence interval:
```{r}
confint(chosen.model)
```

The trend fitted by the model is: 

```{r}
par(mfrow=c(1,1), font = 2, font.axis = 2, font.lab = 4, las = 1, mar = c(0, 0, 1, 0))
plot(chosen.model, se = F, locations = islet_window)
```

Now we assess the goodness of fit of the chosen model:

- via Chi-squared test 

```{r warning=FALSE}
M <- quadrat.test(chosen.model, nx = 4, ny = 2)
M
```

```{R}
par(mfrow=c(1,1), font = 2, font.axis = 2, font.lab = 4, las = 1, mar = c(0, 0, 0, 0))
plot(n84, main="")
plot(M, add = TRUE, cex = 1.5, col = "red")
```

The p-value is very low, even though in almost all the subregions the standardized residuals have absolute value around $2$ or smaller.

- via Kolmogorov-Smirnov test, to check the fitting of the model by each covariate separately.

```{r}
KS1=cdf.test(chosen.model,"y")
KS1
```


```{r}
KS2=cdf.test(chosen.model,Vegetation)
KS2
```

```{r}
KS3=cdf.test(chosen.model,Height)
KS3
```

The $y$ variable seems to fits quite well the data, while Vegetation and Height have very little significance.

- looking at the Smoothed Pearson residuals
```{r}
band=40
par(mfrow=c(1,1), font = 2, font.axis = 2, font.lab = 4, las = 1, mar = c(0, 0, 1, 0))
smooth = diagnose.ppm(chosen.model, which = "smooth", type="pearson", sigma=band)
smooth
```

The standard deviation of the smoothed residuals under the null hypothesis of correct fitting is 
```{r}
smooth$smooth$sdp
```

so residuals whose modulus is larger than this value can be considered as extreme. This means that almost all subregions are not well fitted.

- through a Lurking plot

```{r}
lurking(chosen.model, Height, type = "raw")
```

The residuals lie outside the envelope. The fit is not good:

- When the height is lower than 10 the residuals are consistently negative. Therefore, there are less points than expected with low height.
- From $height=10$ to $height=25$ the residuals are growing, so there are more points than expected.

After conducting these analyses, we can conclude that the model we have built, although it had the best fit from a statistical standpoint, does not actually fit the data well.
Despite this, it was the most significant model based on the data we have.


\newpage

# Task 3: Interaction

### The nest data from islet “nucli 84” is stored in nucli84.txt. Additionally, the coordinates of the islet are in poly84.txt

### 1. Build a ppp object using the “nucli 84” data.

```{r include=FALSE}
library(spatstat)
```

```{r}
rm(list=ls())
nucli84 <- read.delim("data/T3/nucli84.txt")

min.X = min(nucli84$X)
min.Y = min(nucli84$Y)
max.X = max(nucli84$X)
max.Y = max(nucli84$Y)

nucli84$X = nucli84$X - min.X
nucli84$Y = nucli84$Y - min.Y

n84=ppp(x = nucli84$X, y = nucli84$Y, range(nucli84$X), range(nucli84$Y))

poligon=read.delim("data/T3/poly84.txt")

poligon$X = poligon$X - min.X
poligon$Y = poligon$Y - min.Y
pol.illa<-list(x=poligon$X,y=poligon$Y)

min.pX = min(poligon$X)
min.pY = min(poligon$Y)
max.pX = max(poligon$X)
max.pY = max(poligon$Y)

n84p=ppp(nucli84$X,nucli84$Y,poly=pol.illa, range(poligon$X), range(poligon$Y))
```


```{r}
par(mfrow=c(1,1), font = 2, font.axis = 2, font.lab = 4, las = 1, mar = c(0, 0, 2, 0))
plot(n84p,main="Nests")
axis(1,at=c(round(seq(min.pX, max.pX,length=10),digits=0)),
     pos=c(0,0))
axis(2,at=c(round(seq(min.pY, max.pY,length=10),digits=0)),pos=c(min.pX-4,min.pY-50))

```

### 2) Check the interaction pattern.

To check the interaction pattern, we utilize the L-function defined as $L(r) = \sqrt{\frac{K(r)}{\pi}}$ as it is an estimator with a variance that does not vary with respect to $r$. This makes it more stable than other estimators, such as F, G, and K.

```{r}
E <- envelope(n84p, Lest, nsim = 19, rank = 1, global = TRUE,correction="best")
```

```{r}
plot(E, main = "Global envelopes of L(r)", legend=F)
abline(v=1, lty=3)
```

The observed $L(r)$ goes above the theoretical function, indicating that the observed distance between locations is greater than expected under the assumption of a homogeneous Poisson process. This suggests the presence of a regular pattern. Additionally, we notice that the observed line falls outside of the envelopes from $r=1$ onward.

### 3) Model the relation between the intensity of the point process and the covariates height and vegetation accounting for the interaction pattern. Interpret the estimates.

Since we have a regular pattern, we will use Gibbs models.

First we try to fit a Hardcore model:

```{r}
grid <- read.delim("data/T2/grid.txt")
grid.veg = read.delim("data/T2/grid_veg.txt")
veg = as.matrix(read.delim("data/T2/veg.txt",header=FALSE))
height = as.matrix(read.delim("data/T2/height.txt", header=FALSE))

Height = im(mat = height, xcol = grid$x, yrow = grid$y)
Vegetation = im(mat = veg, xcol = grid.veg$x, yrow = grid.veg$y)

fitH<-ppm(n84p, ~ Height + Vegetation, Hardcore)
```

```{r}
summary(fitH)
```

We assess the goodness of fit of the model by looking at the theoretical envelopes of $L(r)$ and comparing it with the observed values.

```{r}
E <- envelope(fitH, Lest, nsim = 19, rank = 1, global = TRUE, correction="best")
```

```{r}
plot(E, main = "global envelopes of L(r)", legend=F)
```

The observed function falls outside of the envelopes, indicating that the Hardcore model does not fit the data well. 

As an alternative, we will try using a Strauss model. Our exploratory analysis showed that the observed L-function lies outside the envelope for values of $r$ greater than 1. To determine the optimal value for $r$, we will analyze the profile likelihood using a range of values inside the interval $[1,15]$ and choose the one that provides the best fit to the data.

```{r}
df=data.frame(r=seq(1,15,by=0.5))
pfit=profilepl(df,Strauss,n84p,~Height + Vegetation)
```

```{r}
plot(pfit, main="Profiling of the Strauss model through logLikelihood")
```

```{r}
r.opt = pfit$fit$fitin$interaction$par$r
```

The optimal value is reached at $r=$ `r r.opt`.

So now we can finally estimate the model:

```{r}
fitSt<-ppm(n84p, ~Height + Vegetation, Strauss(r=r.opt))
fitSt
```

and assess his validity:

```{r}
E <- envelope(fitSt, Lest, nsim = 19, rank = 1, global = TRUE,correction="best")
plot(E, main = "global envelopes of L(r)")
```

The observed L-function is inside the envelopes, therefore the Strauss model can be valid and we proceed to interpret it.

```{r}
summary(fitSt, fine = F)
```

The estimates of its coefficients are:

```{r}
fitSt$coef
```

The estimate for the interaction parameter is	$\gamma =$ `r fitSt$coef[4]`. 
The parameter $\gamma$ controls the strength of interaction between points. If $\gamma = 1$ the model
reduces to a Poisson process. If $\gamma = 0$ the model is a hard core process. For values $0 < \gamma < 1$, like in our case, the process exhibits a moderate strength of interaction between the presence of points in different locations.


\newpage

# TASK 4: Case-control studies

```{r include=FALSE}
# knitr::opts_chunk$set( echo=F )
knitr::opts_chunk$set( cache=T )
knitr::opts_chunk$set(out.width = '70%')
knitr::opts_chunk$set(fig.align = "center")
knitr::opts_chunk$set(tidy =TRUE)
knitr::opts_chunk$set(message = FALSE)
```

### Using the Primary Biliary Cirrhosis data, perform a case-control point pattern analysis. Point locations and marks are stored in *PBCdata.txt” file. Coordinates of the window are in “PBCpoly.txt” file.

The data are:

```{r}
library(spatstat)
rm(list=ls())

pbc.data <- read.delim("data/T4/PBCdata2.txt")
head(pbc.data)
```


### 1) Create the ppp object.

```{r}
poligon=read.delim("data/T4/PBCpoly.txt")
min.pX = min(poligon$x)
max.pX = max(poligon$x)
min.pY = min(poligon$y)
max.pY = max(poligon$y)

pol.illa<-list(x=poligon$x, y=poligon$y)

pbc.data$marks = factor(pbc.data$marks)

pbc.p = ppp(pbc.data$x,pbc.data$y, poly=pol.illa, range(poligon$x), range(poligon$y), marks = pbc.data$marks)
```

```{r}
cases = split(pbc.p, pbc.data$marks)$case
controls = split(pbc.p, pbc.data$marks)$control

par(mfrow=c(1,1), font = 2, font.axis = 2, font.lab = 4, las = 1, mar = c(0, 0, 2.5, 0))
plot(pbc.p$window, main = "Primary Biliary Cirrhosis data")
points(controls, pch = 3, col = "green")
points(cases, pch = 16, col = "red")
legend("right", c("control", "case"), pch = c(3, 16), col=c("green", "red"), bty = "n", cex = 1.2)
```

### 2) Assess the spatial variation of the risk

```{r include=FALSE}
library(sparr)
```

We plot the relative risk in logarithmic scale, so the locations colored in red have a Relative risk greater than 1, while locations colored in blue have a Relative risk smaller than 1.

```{r}
chp <- risk(cases,controls, adapt = T)

M = max(chp$rr)
m = min(chp$rr)

par(mfrow=c(1,1), font = 2, font.axis = 2, font.lab = 4, las = 1, mar = c(0, 0, 1, 0))
plot(chp$rr, gamma = 1.3, main="Relative risk", col=beachcolours(c(m,M)))
```

Let’s use a permutation test to see if there is a significant difference between cases and controls risk, computing on a regular grid the statistic:

$$\hat{T} = |c|\sum_{i=1}^p(\hat{\rho}(s_i) - \hat{\rho}_0)^2$$

where $c$ is the cell of the grid and (under the null hypothesis $\text{H}_0$) $\rho_0=0$.

```{r}
cellsize<-chp$rr$xstep*chp$rr$xstep
rho0<-0
```

The value of the statistic is:

```{r}
ratiorho <- cellsize*sum((chp$rr$v-rho0)^2,na.rm=T)
ratiorho
```

```{r}
# Permutation function
perm_rr<-function(){
  new_ch<-rlabel(pbc.p)
  
  new_cases <- split(new_ch,  f = new_ch$marks)$case
  new_controls <- split(new_ch, f = new_ch$marks)$control

  new_chp <- risk(new_cases,new_controls)
  cellsize<-new_chp$rr$xstep*new_chp$rr$ystep
  ratio_perm <- cellsize*sum((new_chp$rr$v-rho0)^2,na.rm=T)
  ratio_perm
}
```

```{r include=F, echo=T}
nsim<-99
set.seed(2022)
rperm<-sapply(1:nsim,function(i) perm_rr())
```

And its p-value (obtained through a permutation approach) is:

```{r}
(sum(rperm > ratiorho)+1)/(nsim+1)
```

So we don't have enough evidence to reject the null hypothesis.

### 3) Compare the interaction patterns.

We draw the K-functions for cases and controls:

```{r}
s=seq(0,500,length=20)
khcases<-Kest(cases, r=s, correction="best")
khcontrols<-Kest(controls, r=s, correction="best")
```

```{r}
plot(khcases, legend=F)
lines(khcontrols$r,khcontrols$iso,lty=1, col="blue")
legend("topleft", legend = c("cases", "controls", "Poisson"), lty=c(1,1,3), col = c("black", "blue", "red"), lwd=3)
```

The K-functions for both cases and controls are above the Poisson curve, indicating a cluster interaction. To test the null hypothesis that the two K-functions are equal, we will use a permutational approach.

```{r}
Kdif<-function(X, r, cr="iso")
{
  k1<-Kest(X[marks(X)=="case"], r=r, correction=cr)
  k2<-Kest(X[marks(X)=="control"], r=r, correction=cr)
  D=k1[[cr]]-k2[[cr]]
  res<-data.frame(r=r, D=D)
  return(fv(res, valu="D", fname="D"))
}

nsim<-39
envKdif<-envelope(pbc.p, Kdif, r=s ,nsim=nsim, savefuns=TRUE,simulate=expression(rlabel(pbc.p)))
```

```{r}
plot(envKdif,legend=F, main="Envelope of the difference")
```

As shown in the plot, the difference between the K-functions falls outside the envelopes for all ranges, indicating that the difference is not significant. The p-value for this test is as follows:

```{r}
simfuns<-as.data.frame(attr(envKdif, "simfuns"))[,-1]
khcovdiag<-apply(simfuns, 1, var)
T0<-sum( ((khcases$iso-khcontrols$iso)/sqrt(khcovdiag))[-1])
T_pm<-apply(simfuns, 2, function(X){
  sum((X/sqrt(khcovdiag))[-1])
})
pvalue<-2*(sum(abs(T_pm)>abs(T0))+1)/(nsim+1)
pvalue
```

Therefore, we can reject the null hypothesis at a significance level of 95% and conclude that the interaction pattern differs between cases and controls.

